### 2017-01-17 doc/manuscript.Rmd
Re-arranged and added to discussion - prev lit and interpret

TAGFA can fluctuate substantially throughout the day and to control for this all
PROMISE participants came for their clinic visits in the morning in a fasted
state. Some of the inherent variability in TAGFA fluctuation can be managed by
standardizing the data collection time, though there may still be some
{{residual error? random error?}}.

@@

- previous research has shown that CHO intake increases DNL {{ cite }}. In
particular, DNL from CHO sources increases especially the 14-16 chain TAGFA as
well as the 18 chain TAGFA. Prev lit has also shown that higher DNL can increase
risk for DM {{ cite }} and that higher refined CHO can increase risk of DM {{
cite }}. Our study found that the 14-16 TAGFA all strongly predicted lower
insulin sensitivity, reinforcing this finding. While these FA also had a
significant association with beta-cell function, the results were not as strong
as with the IR. {{ why is it that DNL increases risk? }}
- TAGFA that tended to positively associated with IR and BCF also tended to be
the very long chain FA such as eicosapentaenoic acid (20:5n-3) and
docosahexaenoic acid (22:6n-3), which have potent anti-inflammatory properties
{{ cite }}. Biologically {{ why these FA and how help with IR }}
- {{ compare models with and without NEFA }} NEFA have been shown contribute
substantially to TAGFA composition and concentration. Adjusting for ...
- Dietary FA have been shown to contribute approx 25% of the newly synth
VLDL-TAG {{ cite }}. This suggests that our results may not entirely reflect
diet, but also adipose stores. {{ discuss results of NEFA paper }} However, in
an analysis of the NEFA pool {{ cite }}, we found that total NEFA, but not the
specific composition associated with beta-cell function. There may be something
unique about the specific composition of the TAGFA fraction that may allow it to
enter IR tissues more easily than NEFA. It may be that VLDL receptors allow FA
to transition to IR tissue compared to NEFA bound to albumin {{ confirm and cite
}}.
- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
- Differences with 
- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
lipotoxic environment {{ cite }}

[@Barrows2006a] NEFA contributes 60-80% of newly synth VLDL-TAG during fasting
and ~40% during fed. Dietary FA contribute ~25% to newly synth VLDL-TAG.
VLDL-TAG is not cleared (lipolyzed) as much as chylomicron and so stays in
circulation longer (hence why there is bigger TAG pool vs NEFA)

Previous lit and limits:
- A few epidemiological studies have examined the role of the TAGFA composition
on insulin resistance and diabetes risk. 
- A weight loss intervention trial showed that changes in the TAGFA composition
of 19 participants accompanied changes in the weight and insulin sensitivity over
the 33 week period [@Schwab2008a]. Specifically, the TAGFA 14:0, 14:1, 16:0,
16:1, 18:0, 18:3n-3, and 18:3n-6 were all lower after weight loss and correlated
with improved insulin sensitivity.
- In a short (24hr) feeding trial of 24 subjects [@Lee2015a], palmitoleic acid
(16:1n-7) was shown to be strongly correlated with isotopically measured DNL and
was 45% higher in individuals with a greater amount of hepatic fat. However, in
that study only the TAGFA 18:0 was correlated with HOMA-IR, likely due to the
small sample size.
- To our knowledge, there have been two prospective studies that had a similar
research objective as our cohort. One was a nested case-control cohort of the 
Framingham offspring study [@Rhee2015a], which found that TAGFA with a lower
carbon chain and less double bonds (e.g. 14:0, 16:0) had higher OR for diabetes
after 12-years while higher carbon chain and more double bond TAGFA had a lower
OR for diabetes. This pattern of TAGFA with diabetes risk remained the same when
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
TAGFA data on 831 participants after 6-years from the baseline visit. This
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. A beta-cell function measure was computed, but only included 
calculations up to 30 minutes of the OGTT. As with our results, they found a
larger effect size for insulin sensitivity compared to the beta-cell function.
The specific TAGFA that associated with the metabolic measures were similar 
between studies, except they did not find 
- The primary limit to these studies is the mainly small sample size and short
study duration. Two studies had larger sample sizes of 831 [@Lankinen2015a] and 
`r 189+189` [@Rhee2011a] and both were prospective cohorts. In addition, the
Framingham offspring study [@Rhee2011a] did not have data on beta-cell function 
while the Finnish cohort [@Lankinen2015a] used {{ how do I say this politely Tony? 
They're methods and article was ... questionable... }} simple statistical 
techniques to analysis the longitudinal data, which do not adequately take into
consideration auto-correlation of longitudinal values and the high dimensionality
of the TAGFA composition data structure.

- Refs:
    - Lankinen2015a (poor study, but is unfortunately relevant...) (god... I
    really don't want to cite/talk about this...)
    - Lee2015a (16:1n7 as marker of DNL) (n=24, feeding trial, 16:1n7 was 45%
    higher in high liver fat group, direct dnl (isotope) was 2.5 fold higher in
    high liver fat, 16:1n7 was corr (0.56) with direct DNL, TAG 18:0 corr (.50)
    with HOMA)
    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
    two groups: normogly and hypergly. results in general showed no diff between
    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
    not present results on whole fraction for IR)
    - Rhee2011a (good paper on TAGFA composition) (Offspring of Framingham...,
    nested case-control, n=189 case, n=189 control, MS/MS, OGTT, differences in
    fasting TAGFA were more apparent than at 2-hr ogtt, more lower chain/lower
    desat in cases + higher OR, while higher chain/desat had lower OR, lower
    chain/desat pos corr with HOMA-IR, MS/MS (tandem) showed that 12-18 SFA and
    MUFA inc risk of diabetes)
    - Kotronen2009a (TAGFA composition) (n=16, 18-60, mostly female, lipidomics,
    GC, HOMA-IR, total esterified (of which TAGFA makes up the majority) 16,
    161n7, 181n9 pos corr with homa, 182n6 neg with homa)
    - Schwab2008a (TAGFA composition and insulin sensitivity) (n=75 (but only
    n=19 for lipidomics), overweight or obese, weght loss intervention (33
    weeks), weight, FSIGT, GC, UPLC-MS, after weight loss 14, 141, 16, 161, 18,
    183n3, 183n6 were lower, lower TAGFA that contained 14, 14:1, 16, 161, 18,
    181 were correlated with improved Si)

[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
control than 2hr-OGTT TAG
[@Kotronen2009a] (this wasn't studied, but discussed) higher hepatic DNL may
increase amount of 'shorter' chain FA (16:0, 14:0, etc.) into VLDL and hence
into circulation. So higher of these FA may indicate higher DNL -> problems with
liver function -> maybe hepatic IR -> cyclical?
[@Lee2015a] 16:1n-7 may be good marker of liver DNL.

Wilke2009a - contributions to DNL (e.g. 70% 16:0 to newly synth)
The four FA that cluster together on the PLS also appear to be the same FA that
make up most of DNL, especially in context of higher carb/refined carb in diet
[@Wilke2009a].

New things:

- No longitudinal study has examined all these TAGFA using detailed metabolic,
OGTT-derived measures. No study has used these statistical techniques, which
were designed with these types of data in mind, and as such allows appropriate
conclusions to be reached from this data.
- That 14-16 FA in the TAGFA fraction predict lower insulin sensitivity and that
this particular cluster is also generally observed with higher CHO intake.

@@

Since no interaction, no loop? Not enough sample, already IR? maybe?

Concentration data is not

Discuss well done paper (very relevant) Kotronen2009a

>>>>

TAGFA can fluctuate substantially throughout the day and to control for this all
PROMISE participants came for their clinic visits in the morning in a fasted
state. Some of the inherent variability in TAGFA fluctuation can be managed by
standardizing the data collection time, though there may still be some
{{residual error? random error?}}.
[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
control than 2hr-OGTT TAG

@@

Previous research has shown that carbohydrate intake increases DNL {{ cite }}. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for diabetes {{ cite }}
as well as studies showing higher refined carbohydrates (e.g. added sugars) and
a higher risk for diabetes {{cite}}. Our study extends {{adds to? confirms?}}
these findings on DNL by showing that TAGFA with 14 to 16 carbons all strongly
predicted lower insulin sensitivity. While these FA also had a significant
association with beta-cell function, the results were not as strong as with the
IR. {{ why is it that DNL increases risk? }}

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
statistical techniques designed to adequately analyze ... {{shots fired? 
necessary?}}.

To our knowledge, two prospective studies have been published that had similar
research objectives as the present analysis. One was a nested case-control
cohort (n=189 cases and n=189 controls) of the Framingham offspring study
[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and less double
bonds (e.g. 14:0, 16:0) had a higher OR {{confirm}} for diabetes after 12-years
while higher carbon chain and more double bond TAGFA had a lower OR for
diabetes. This pattern of TAGFA with diabetes risk remained
the same when they examined HOMA-IR cross-sectionally at the baseline visit. The
other study was a prospective cohort of males in Finland [@Lankinen2015a], of
which they had TAGFA data on 831 participants after 6-years from the baseline
visit. This cohort has similar outcomes as PROMISE, however, OGTT data was only
available at the 6-year visit. A beta-cell function measure was computed, but
only included calculations up to 30 minutes of the OGTT. As with our results,
they found a larger effect size for insulin sensitivity compared to the
beta-cell function. The specific TAGFA that associated with the metabolic
measures were similar between studies, except they did not find {{confirm}}

In addition, the Framingham offspring study [@Rhee2011a] did not have data on
beta-cell function while the Finnish cohort [@Lankinen2015a] used {{ how do I
say this politely Tony? They're methods and article was ... questionable... }}
simple statistical techniques to analysis the longitudinal data, which do not
adequately take into consideration auto-correlation of longitudinal values and
the high dimensionality of the TAGFA composition data structure.

A smaller study (n=16) of mostly females using highly detailed lipodomic
techniques found a positive correlation between total esterified (of which TAGFA
make up the majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a].

Limitations with estimated DNL or measured DNL in prev studies.

Several studies have been conducted that link greater hepatic DNL with the 14
and 16 carbon FA. A short, small feeding trial (n=24) was conducted to identify
which biomarkers most accurately reflected DNL [@Lee2015a]. The study found that
16:1n-7, directly measured DNL using isotopes, and liver fat were all highly
correlated with each other, suggesting that 16:1n-7 is a good biomarker for
hepatic DNL.
In another small (n=14) feeding trial, meal type (high fat vs low fat) was
tested to determine the effect on DNL and TAGFA composition [@Wilke2009a]. They
found that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together, suggesting {{another word}} that these fatty acids are
indicative of a higher carbohydrate diet.
Suggesting {{expand/synth}} that greater hepatic fat may also contribute to greater DNL,
increasing the 14 and 16 carbon FA in circulation, potentially exposing tissues
to greater lipotoxicity, resulting in further hepatic dysfunction and continuing
the cycle.

- contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}

- A weight loss intervention trial showed that changes in the TAGFA composition
of 19 participants accompanied changes in the weight and insulin sensitivity over
the 33 week period [@Schwab2008a]. Specifically, the TAGFA 14:0, 14:1, 16:0,
16:1, 18:0, 18:3n-3, and 18:3n-6 were all lower after weight loss and correlated
with improved insulin sensitivity.
{{ hard to disentangle reverse causality here }}
    - Schwab2008a (TAGFA composition and insulin sensitivity) (n=75 (but only
    n=19 for lipidomics), overweight or obese, weght loss intervention (33
    weeks), weight, FSIGT, GC, UPLC-MS, after weight loss 14, 141, 16, 161, 18,
    183n3, 183n6 were lower, lower TAGFA that contained 14, 14:1, 16, 161, 18,
    181 were correlated with improved Si)

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL {{cite}}. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
in adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
contributed ~25% to newly synth VLDL-TAG {{fasting or fed?}} [@Barrows2006a].
While the NEFA fraction is much smaller than the TAG fraction even though it
contributes substantially to the TAG fraction, VLDL-TAG is not cleared
(lipolyzed) as much as chylomicron (which also contains a large amount of TAG)
and so has a higher half-life than NEFA, remaining in circulation longer {{hence
why there is bigger TAG pool vs NEFA}}. In an analysis of the NEFA fraction in
the PROMISE cohort with the OGTT-derived outcomes (currently unpublished)
{{correct word for this??}}, we found that total NEFA, but not the specific
composition associated with beta-cell function and not with insulin resistance.
While NEFA may contribute substantially to the TAG fraction, there may be
something unique about the specific composition of the TAGFA fraction that may
allow it to enter insulin sensitive tissues more easily than NEFA. It may be
that VLDL receptors allow FA to transition to these tissue compared to NEFA
bound to albumin {{ confirm and cite }}, and thus able to exert potential
lipotoxic effects. {{ comment on comparing models with and without NEFA }}

@@

As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
that cis-vaccenic acid strongly predicted higher insulin sensitivity and
beta-cell function. Likewise, in the TAGFA fraction we found a similar
association with cis-vaccenic acid. {{ continue? }}
    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
    two groups: normogly and hypergly. results in general showed no diff between
    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
    not present results on whole fraction for IR)

{{ comment on unadjusted vs adjusted }}

{{ commment on null interactions with time }}
- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
Since no interaction, no loop? Not enough sample, already IR? maybe?

- Or that post-prandial TAG is not suppressed in IR Sondergaard2012a
- {{ confirm }} greater flux of saturated fatty acids into tissues may encourage
lipotoxic environment {{ cite }}

{{ Concentration data is not }}
